---
title: "QC"
author: "astridhofman7"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Set-Up

```{r message = FALSE, echo = FALSE, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(Seurat) 
library(hdf5r)
library(BiocManager)
library(limma)
library(sctransform)
library(edgeR)
#library(enrichR)
library(scDblFinder)
#library(Azimuth)
#library(ezRun)
library(patchwork)
library(cowplot)
library(scuttle)
library(Matrix)
library(ggrepel)
library(EnhancedVolcano)
library(ggplot2)
#library(gridExtra)
library(tidyverse)
library(SeuratData)
library(RColorBrewer)
library(Matrix)
library(scDblFinder)
library(workflowr)
library(here)
library(scCustomize)
library(scDblFinder)
library(harmony)
library(clustree)


set.seed(42)

#Output paths
current_file <- "QC"

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)
```

```{r knitr config, include=FALSE, cache=FALSE}

knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 600, fig.align = "center")

```

## Load the samples & set directories

```{r error=TRUE, message = FALSE, include=FALSE, cache=FALSE}
#To load the filtered feature matrices using the here package
##Set the folders/paths to find the necessary files
folders <- c("HC-01", "RES-001", "RES-003", "RES-004")
files <- setNames(
  lapply(folders, function(folder) {
    here("data", folder,"filtered_feature_bc_matrix")
  }),
  folders
)

#Check that the file paths and directory exist
print(files)
dir_exists <- sapply(files, dir.exists)
print(dir_exists)

# Create a list with the seurat  objects
seurat_list <- lapply(names(files), function(sample) {
  data <- Read10X(data.dir = files[[sample]])  # Load data using Read10X
  seurat_object <- CreateSeuratObject(counts = data, project = sample)  # Create Seurat object
  seurat_object$sample <- sample  # Add sample identity metadata
  return(seurat_object)
})

# Name the list by the folder names
names(seurat_list) <- names(files)
```


## Inspect the quality of the data before removing doublets or filtering

```{r}
#Calculate and add the percentage of mitochondrial and ribosomal genes
seurat_list <- lapply(seurat_list, function(seurat_object) {
  seurat_object[["percent.mt"]] <- PercentageFeatureSet(seurat_object, pattern = "^MT-") 
  return(seurat_object)
})
seurat_list <- lapply(seurat_list, function(seurat_object) {
  seurat_object[["percent.rp"]] <- PercentageFeatureSet(seurat_object, pattern = "^RP[SL]") 
  return(seurat_object)
})

seurat_list <- lapply(seurat_list, function(seurat_object) {
  seurat_object[["percent.hb"]] <- PercentageFeatureSet(seurat_object, pattern = "^HB[^(P)]") 
  return(seurat_object)
})

```

```{r, fig.height=4, fig.width=12}

#Visualize QC metrics for each sample in seurat list-Violin Plots
qc_plots <- lapply(seurat_list, function(seurat_object) {
  VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt","percent.rp"), ncol = 4) 
  })

# Print QC plots
for (i in seq_along(qc_plots)) {
  print(qc_plots[[i]])
}

```


```{r, fig.height=4, fig.width=12}
#Visualize QC metrics for each sample in seurat list-Scatter Plots
scatter_plots <- lapply(seurat_list, function(seurat_object) {
  FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "percent.rp") 
})

# Print scatter plots
for (i in seq_along(scatter_plots)) {
  print(scatter_plots[[i]])
}

```

```{r, message=FALSE}
#Merge seurat list in a seurat object
# Find common features (genes) across all Seurat objects in the list
common_features <- Reduce(intersect, lapply(seurat_list, rownames))

# Subset each Seurat object to retain only common features
seurat_list <- lapply(seurat_list, function(obj) {
    obj <- subset(obj, features = common_features)
    return(obj)
})

# Merge the Seurat objects without performing any normalization
merged_seurat <- Reduce(function(x, y) merge(x, y, assay = "RNA"), seurat_list)

merged_seurat$Sample <- merged_seurat$orig.ident
```
## Doublet Detection & Removal

```{r, message=FALSE}
#ScDblFinder to detect doublets
#Convert you seurat object to a single cell experiment for ScDblFinder
#Your Seurat is a v5 object
#Join layers

merged_new = JoinLayers(merged_seurat)

#Extract raw counts 

counts_data <- SeuratObject::GetAssayData(merged_new, slot = "counts")


#Extract metadata (cell-level data)

metadata <- merged_new@meta.data
 
#Extract feature (gene) information if needed

feature_metadata <- data.frame(row.names = rownames(counts_data))

 
sce <- SingleCellExperiment::SingleCellExperiment(

    assays = list(counts = counts_data),  # Add other slots if needed (e.g., "logcounts")

    colData = metadata,                  # Add cell metadata

    rowData = feature_metadata           # Add feature metadata (optional)

)


#Find Doublets

sce <- scDblFinder(sce, samples="Sample", clusters = TRUE)
table(sce@colData$scDblFinder.class)
sce <- logNormCounts(sce)
```

### Convert back to Seurat Object

```{r}
#Convert back to a seurat object
merged_new <- as.Seurat(sce, counts = "counts", data = "logcounts")
rm(sce)
merged_new@meta.data[merged_new@meta.data$scDblFinder.class %in% "singlet", "scDblFinder.n"] = paste0("Singlets (n=", table(merged_new$scDblFinder.class)[1], ")")
merged_new@meta.data[merged_new@meta.data$scDblFinder.class %in% "doublet", "scDblFinder.n"] = paste0("Doublets (n=", table(merged_new$scDblFinder.class)[2], ")")
merged_new$scDblFinder.n <- factor(x = merged_new$scDblFinder.n, levels = c(rownames(table(merged_new$scDblFinder.n))[2], rownames(table(merged_new$scDblFinder.n))[1])) 
```

### Plot the Detected Doublets

```{r fig.width=12}
p1 <- VlnPlot(merged_new, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 150000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(merged_new, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

Doublets <- p1 + p2 + plot_layout(ncol = 2)
Doublets
```

### Check the QC Parameters before Removing Doublets for each sample

```{r}
#Check quality control parameters before removing doublets by sample
#Use the seurat object with joined layers
p1 <- VlnPlot(merged_new, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 40000) + theme(axis.title.x = element_blank()) + NoLegend()
p2 <- VlnPlot(merged_new, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 6000) + theme(axis.title.x = element_blank()) + NoLegend()
p3 <- VlnPlot(merged_new, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()
p4 <- VlnPlot(merged_new, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

before_removal <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
before_removal
```

### Doublet Removal

```{r}
#Remove Doublets
nodoublets <- subset(x = merged_new, subset = scDblFinder.class == "singlet")
table(nodoublets@meta.data$scDblFinder.class)
```

```{r}
#Check quality control parameters after removing doublets by sample

p1 <- VlnPlot(nodoublets, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 40000) + theme(axis.title.x = element_blank()) + NoLegend()
p2 <- VlnPlot(nodoublets, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 6000) + theme(axis.title.x = element_blank()) + NoLegend()
p3 <- VlnPlot(nodoublets, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()
p4 <- VlnPlot(nodoublets, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

after_removal <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
after_removal
```
## Checking the Genes & Percent MT and how these correlate
```{r}
nodoublets@meta.data %>% 
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point(alpha = 0.6) + 
  scale_color_gradient(low = "lightgreen", high = "darkgreen") + 
  stat_smooth(method = "lm", aes(group = Sample), color = "black") +  # Add `group = Group` to smooth
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  xlab('Number of UMIs (n_Counts)') + 
  ylab('Number of Genes (n_Feature)') +
  theme(axis.text.x=element_text(size=13), axis.text.y=element_text(size=15), 
        axis.title = element_text(size = 22), 
        plot.title = element_text(size = 24), 
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 18)) +
  facet_wrap(~Sample)
```

```{r}
nodoublets@meta.data %>% 
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point(alpha = 0.6) + 
  scale_color_gradient(low = "lightblue", high = "darkblue") + 
  stat_smooth(method = "lm", color = "black") +  # Add `group = Group` to smooth
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  xlab('Number of UMIs (n_Counts)') + 
  ylab('Number of Genes (n_Feature)') +
  theme(axis.text.x=element_text(size=13), axis.text.y=element_text(size=15), 
        axis.title = element_text(size = 22), 
        plot.title = element_text(size = 24), 
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 18))
```

## Filtering 

```{r}
#Filtering based on number of counts, features and mitochondrial percentage
# Split Seurat object by sample
seurat_list <- SplitObject(nodoublets, split.by = "sample")
quantiles_list <- list()

# Define the list of sample names
samples <- unique(nodoublets@meta.data$sample)

# Split the object by "Sample"
seurat_list <- SplitObject(nodoublets, split.by = "sample")

# Initialize a list to store quantile information
quantiles_list <- list()

# Loop through each sample
for (sample in samples) {
  seu1 <- seurat_list[[sample]]
  
  # Calculate quantiles for the current sample
  quantile_df <- data.frame(
    percent_mt_q90 = quantile(seu1@meta.data$percent.mt, probs = 0.90),
    percent_mt_q10 = quantile(seu1@meta.data$percent.mt, probs = 0.10),
    nCounts_q90 = quantile(seu1@meta.data$nCount_RNA, probs = 0.90),
    nCounts_q10 = quantile(seu1@meta.data$nCount_RNA, probs = 0.10),
    nFeature_q90 = quantile(seu1@meta.data$nFeature_RNA, probs = 0.90),
    nFeature_q10 = quantile(seu1@meta.data$nFeature_RNA, probs = 0.10),
    Ncell_before_filtering = ncol(seu1)
  )
  
  # Store the quantiles for the current sample
  quantiles_list[[sample]] <- quantile_df
}


# Combine the quantiles list into a single data frame
quantiles_df <- bind_rows(quantiles_list, .id = "Sample")

# Gather measures into a long format for easier plotting
quantiles_long <- quantiles_df %>%
  pivot_longer(cols = -Sample, 
               names_to = "Metric", 
               values_to = "Value")

# Function to plot violins for each measure
generate_violin_plots <- function(seurat_list, measures) {
  plots <- list()
  for (measure in measures) {
    # Combine meta.data from all Seurat objects into one data frame
    all_data <- do.call(rbind, lapply(names(seurat_list), function(sample) {
      data.frame(Sample = sample, Value = seurat_list[[sample]]@meta.data[[measure]])
    }))
    
    # Generate the violin plot
    p <- ggplot(all_data, aes(x = Sample, y = Value, fill = Sample)) +
      geom_violin(alpha = 0.6) +
      geom_hline(data = quantiles_long %>%
                   filter(grepl(measure, Metric)), 
                 aes(yintercept = Value, linetype = Metric), 
                 color = "red") +
      scale_linetype_manual(values = c("dashed", "solid")) +
      labs(title = paste("Violin Plot of", measure),
           x = "Sample", y = measure) +
      theme_minimal()
    
    # Save the plot in the list
    plots[[measure]] <- p
  }
  return(plots)
}

# Define the measures to plot
measures <- c("percent.mt", "nCount_RNA", "nFeature_RNA")

# Generate plots
violin_plots <- generate_violin_plots(seurat_list, measures)

# Print all plots
for (measure in names(violin_plots)) {
  print(violin_plots[[measure]])
}


```

```{r}
rm(quantile_df)
 
# Initialize a list to store the filtered Seurat objects
filtered_seu_list <- list()
 
# Loop through each sample and apply the filtering criteria
for (sample in samples) {
 
  # Extract the corresponding quantiles for the sample
  sample_quantiles <- quantiles_list[[sample]]
  seu1 <- seurat_list[[sample]]
  # Apply the filtering criteria
  seu1 <- subset(seu1, subset = nFeature_RNA > sample_quantiles$nFeature_q10 &
                              nFeature_RNA < sample_quantiles$nFeature_q90 &
                              nFeature_RNA > 300  &
                              percent.mt < sample_quantiles$percent_mt_q90 &
                              percent.mt < 15 &
                              percent.mt > 0.5 &
                              nCount_RNA > 500 &
                              nCount_RNA > sample_quantiles$nCounts_q10 &
                              nCount_RNA < sample_quantiles$nCounts_q90)
 
  # Add the filtered subset to the list
  filtered_seu_list[[sample]] <- seu1
}
```

```{r}
#Merge filtered list to a seurat object
filtered_seurat <- Reduce(function(x, y) merge(x, y, assay = "RNA"), filtered_seu_list)
```

```{r}
#Check quality control parameters after filtering

p1 <- VlnPlot(filtered_seurat, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 15000) + theme(axis.title.x = element_blank()) + NoLegend()
p2 <- VlnPlot(filtered_seurat, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 6000) + theme(axis.title.x = element_blank()) + NoLegend()
p3 <- VlnPlot(filtered_seurat, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()
p4 <- VlnPlot(filtered_seurat, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

filtered_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
filtered_VlnPlots
```

```{r}
#Print a table with filtered samples
table(filtered_seurat$Sample)
```

```{r}
filtered_seurat@meta.data %>% 
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point(alpha = 0.6) + 
  scale_color_gradient(low = "lightgreen", high = "darkgreen") + 
  stat_smooth(method = "lm", color = "black") +  # Add `group = Group` to smooth
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  xlab('Number of UMIs (n_Counts)') + 
  ylab('Number of Genes (n_Feature)') +
  theme(axis.text.x=element_text(size=13), axis.text.y=element_text(size=15), 
        axis.title = element_text(size = 22), 
        plot.title = element_text(size = 24), 
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 18))
```

```{r}
#Filtering bad cells

filtered_seurat <- subset(filtered_seurat, features = rownames(filtered_seurat)[rowSums(filtered_seurat[["originalexp"]]@counts > 0) >= 5])
```

```{r}
#Check quality control parameters after filtering

p1 <- VlnPlot(filtered_seurat, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 15000) + theme(axis.title.x = element_blank()) + NoLegend()
p2 <- VlnPlot(filtered_seurat, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 6000) + theme(axis.title.x = element_blank()) + NoLegend()
p3 <- VlnPlot(filtered_seurat, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()
p4 <- VlnPlot(filtered_seurat, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

filtered_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
filtered_VlnPlots
```

```{r}
#Print a table with filtered samples
table(filtered_seurat$Sample)
```

## Cell Cycle Genes

```{r}
#RES-004 still looks like it has quite some cells with low nCount and nFeature, which could be indicative of low quality cells. Let's take a closer look at some factors that could affect this
# Use built-in cell cycle markers for human
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Apply cell cycle scoring to the Seurat object
filtered_seurat <- CellCycleScoring(filtered_seurat, 
                               s.features = s.genes, 
                               g2m.features = g2m.genes, 
                               set.ident = TRUE)

# Visualize the cell cycle scores
VlnPlot(filtered_seurat, features = c("S.Score", "G2M.Score"), group.by = "Sample", pt.size= 0)

#Conclusion: doesn't look like the cell cycle genes are significantly affecting the expression.
#Continue with the normalization, etc. to see if we can check expression of dying cells
```

## Post-Filtering 

### Normalization

```{r}
#Normalization
filtered_seurat <- NormalizeData(filtered_seurat, normalization.method = "LogNormalize") 
gene_expression <- apply(LayerData(filtered_seurat,layer="data"),1,mean)

names(gene_expression) <- rownames(data)
gene_expression <- sort(gene_expression, decreasing = TRUE) 
head(gene_expression, n=50)
```

### Find Variable Features 

```{r}
#Find Variable Features
FindVariableFeatures(filtered_seurat) -> filtered_seurat
HVFInfo(filtered_seurat) %>%
  arrange(desc(variance.standardized)) %>%
  head(n=20)
HVFInfo(filtered_seurat) %>%
  arrange(desc(variance.standardized)) %>%
  mutate(index=1:n()) %>%
  ggplot(aes(x=index, y=variance.standardized)) +
  geom_line() +
  scale_x_log10() +
  geom_vline(xintercept = 500, colour="red2")

FindVariableFeatures(
  filtered_seurat, 
  selection.method = "vst", 
  nfeatures=500
) -> filtered_seurat
filtered_seurat <- FindVariableFeatures(filtered_seurat, selection.method = "mean.var.plot", nfeatures = 2000)
filtered_seurat <- FindVariableFeatures(filtered_seurat, selection.method = "dispersion", nfeatures = 2000)
top_variable_features <- head(VariableFeatures(filtered_seurat), 10)
print(top_variable_features)

```

### Scaling, PCA, Elbow Plot

```{r}
#Scale data
filtered_seurat <- ScaleData(filtered_seurat, vars.to.regress = "percent.mt")

#PCA
filtered_seurat <- RunPCA(filtered_seurat, npcs = 50, verbose = FALSE)

#Elbow Plot
ElbowPlot(filtered_seurat, ndims=50)

#Choose 25 PCs
```

## Clustering without integration first 

```{r, message=F, echo=F, include=FALSE}
filtered_seurat <- RunUMAP(filtered_seurat, reduction = "pca", dims = 1:25)
filtered_seurat <- FindNeighbors(filtered_seurat, reduction = "pca", dims = 1:25)
resolution.range <- seq(from = 0, to = 0.8, by = 0.1)
filtered_seurat <- FindClusters(object = filtered_seurat, resolution = resolution.range)
```


```{r, fig.width=8, fig.height=15}
#Clustree
clustree(filtered_seurat)
```

### Dimplot / Resolutions

```{r, fig.width=12, fig.height=15}
DimPlot(filtered_seurat, reduction = "umap", shuffle = T,
        group.by = c("originalexp_snn_res.0.1", "originalexp_snn_res.0.2", "originalexp_snn_res.0.3", "originalexp_snn_res.0.4"), ncol = 2)

```

### Dimplot / Sample

```{r, fig.width=10, fig.height=6}

DimPlot(filtered_seurat, reduction = "umap", shuffle = T,
        group.by = c("sample"))

#Clearly needs to be integrated 
```

### Feature plots: Mitochondrial, Ribosomal, Hemoglobin

```{r, fig.width=10, fig.height=5}

# hemoglobin genes (but not HBP)
filtered_seurat <- Seurat::PercentageFeatureSet(filtered_seurat,
                                    pattern = "^HB[^(P)]",
                                    col.name = "percent.hb")
FeaturePlot_scCustom(filtered_seurat, reduction = "umap", features = c("percent.mt", "percent.rp", "percent.hb"))

#No red flags here
```

## Clustering with Integration

### Harmony Integration

```{r}
filtered_seurat[["originalexp"]] <- split(filtered_seurat[["originalexp"]], f = filtered_seurat$sample)

filtered_seurat_harmony <- IntegrateLayers(
  object = filtered_seurat, method = HarmonyIntegration, npcs = 25,
  orig.reduction = "pca", new.reduction = "harmony", max.inter.harmony = 10,
  verbose = TRUE
)
```

```{r, message=FALSE}
filtered_seurat_harmony <- RunUMAP(filtered_seurat_harmony, reduction = "harmony", dims = 1:25, reduction.name = "umap.harm")
filtered_seurat_harmony <- FindNeighbors(filtered_seurat_harmony, reduction = "harmony", dims = 1:25)
resolution.range <- seq(from = 0, to = 0.8, by = 0.1)
filtered_seurat_harmony <- FindClusters(object = filtered_seurat_harmony, resolution = resolution.range)
clustree(filtered_seurat_harmony)
```

```{r, fig.width=18, fig.height=20}

p1 <- DimPlot(
  filtered_seurat_harmony,
  reduction = "umap.harm",
  group.by = c("originalexp_snn_res.0.1", "originalexp_snn_res.0.2", "originalexp_snn_res.0.3", "originalexp_snn_res.0.4", "originalexp_snn_res.0.5", "originalexp_snn_res.0.6"),
  combine = FALSE, label.size = 2
)
wrap_plots(p1, ncol = 3, byrow = F)
```
```{r, fig.width=10, fig.height=5}

DimPlot(
  filtered_seurat_harmony,
  reduction = "umap.harm",
  group.by = c("sample"),
  combine = FALSE, label.size = 2)

```

```{r, fig.width=15, fig.height=5}

DimPlot(
  filtered_seurat_harmony,
  reduction = "umap.harm",
  split.by = c("sample"),
  combine = FALSE, label.size = 2)

```

## Set the resolution 
```{r}
#Will set at 0.6 for now and revisit later in case it doesn't make sense
filtered_seurat_harmony$seurat_clusters <- filtered_seurat_harmony$originalexp_snn_res.0.5
filtered_seurat_harmony <- Seurat::SetIdent(filtered_seurat_harmony, value = filtered_seurat_harmony$originalexp_snn_res.0.5)
table(filtered_seurat_harmony$seurat_clusters)
```

```{r, fig.width= 7, fig.height= 7}

DimPlot(
  filtered_seurat_harmony,
  reduction = "umap.harm",
  label = T, label.size = 5, repel = T)

```
#### Rejoin Layers
```{r}

filtered_seurat_harmony <- JoinLayers(filtered_seurat_harmony)
```

## Annotation (https://www.nature.com/articles/s41467-021-24110-y/figures/1)

### Markers
```{r}

Endo <- c("SELE", "PECAM1","VWF")
Fib <- c("COL1A1", "COL1A2", "COL3A1")
SMC <- c("TAGLN", "ACTA2", "TPM2")
KRT <- c("KRT14", "KRT1", "KRT10", "KRT5")
IMM <- c("LYZ", "HLA-DRA")
LYME <- c("COL21", "LYVE1")
NEU <- c("NRXN1")
SGC <- c("SCGB1B2P", "SCGB1D2")
Mela <- c("TYRP1", "PMEL")
Tcell <- c("CCR7", "CD8A", "IL7R", "CD3E")
```

#### Endothelial Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = Endo, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = Endo, reduction = "umap.harm")
```


```{r, fig.width=10, fig.height=10}
#VlnPlot_scCustom(filtered_seurat_harmony, features = Fib, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = Fib, reduction = "umap.harm")
```

```{r, fig.width=10, fig.height=10}
#VlnPlot_scCustom(filtered_seurat_harmony, features = SMC, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = SMC, reduction = "umap.harm")
```

```{r, fig.width=7, fig.height=7}
#VlnPlot_scCustom(filtered_seurat_harmony, features = KRT, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = KRT, reduction = "umap.harm")
```
#### Immune Cells 

```{r, fig.width=8, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = IMM, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = IMM, reduction = "umap.harm")
```
#### Lymphatic Endothelial Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = LYME, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = LYME, reduction = "umap.harm")
```

#### Neuronal Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = NEU, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = NEU, reduction = "umap.harm")
```
#### Sweat Gland Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = SGC, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = SGC, reduction = "umap.harm")
```

#### Melanocytes

```{r, fig.width=10, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = Mela, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = Mela, reduction = "umap.harm")
```
#### T Cells 

```{r, fig.width=7, fig.height=7}
#VlnPlot_scCustom(filtered_seurat_harmony, features = Tcell, pt.size = 0)
FeaturePlot_scCustom(filtered_seurat_harmony, features = Tcell, reduction = "umap.harm")
```


### What is cluster 17 

```{r, fig.width=7, fig.height=7}
# Create a vector of colors for all clusters
# Use "gray" for all clusters except "17", which will be "red"
all_clusters <- levels(Idents(filtered_seurat_harmony))
cluster_colors <- ifelse(all_clusters == "17", "red", "gray")

# Generate the UMAP plot
DimPlot(filtered_seurat_harmony, 
        reduction = "umap", 
        group.by = "seurat_clusters", 
        label = TRUE) +
  scale_color_manual(values = setNames(cluster_colors, all_clusters)) +
  ggtitle("Highlighting Cluster 17") +
  theme(legend.position = "none")


```


```{r, fig.width=7, fig.height=7}

# Assuming 'seurat_object' is your Seurat object
# Find markers for cluster 17
cluster_17_markers <- FindMarkers(filtered_seurat_harmony, ident.1 = "17")

# Check the number of markers
n_markers_cluster_17 <- nrow(cluster_17_markers)
print(paste("Number of markers for cluster 17:", n_markers_cluster_17))


```

### Remove Cluster 17

```{r, fig.width=7, fig.height=7}

clusters_to_keep <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "18")
Idents(filtered_seurat_harmony) <- filtered_seurat_harmony$originalexp_snn_res.0.5

# Subset the Seurat object based on the clusters
filtered_seurat_harmony <- subset(filtered_seurat_harmony, idents = clusters_to_keep)
table(Idents(filtered_seurat_harmony))

DimPlot(filtered_seurat_harmony, reduction = "umap.harm", label = T, label.size = 5)
table(filtered_seurat_harmony$sample)
```
## Reclustering the UMAP without Cluster 17 


```{r, message = F, echo = F}
#Normalization
skin_final <- NormalizeData(filtered_seurat_harmony, normalization.method = "LogNormalize") 
gene_expression <- apply(LayerData(skin_final,layer="data"),1,mean)

names(gene_expression) <- rownames(data)
gene_expression <- sort(gene_expression, decreasing = TRUE) 
head(gene_expression, n=50)
```

### Find Variable Features 

```{r,  message = F, echo = F}
#Find Variable Features
FindVariableFeatures(skin_final) -> skin_final
HVFInfo(skin_final) %>%
  arrange(desc(variance.standardized)) %>%
  head(n=20)
HVFInfo(skin_final) %>%
  arrange(desc(variance.standardized)) %>%
  mutate(index=1:n()) %>%
  ggplot(aes(x=index, y=variance.standardized)) +
  geom_line() +
  scale_x_log10() +
  geom_vline(xintercept = 500, colour="red2")

FindVariableFeatures(
  skin_final, 
  selection.method = "vst", 
  nfeatures=500
) -> skin_final
skin_final <- FindVariableFeatures(skin_final, selection.method = "mean.var.plot", nfeatures = 2000)
skin_final <- FindVariableFeatures(skin_final, selection.method = "dispersion", nfeatures = 2000)
top_variable_features <- head(VariableFeatures(skin_final), 10)
print(top_variable_features)

```

```{r, fig.width=7, fig.height=7, message = F}
#Scale data
skin_final <- ScaleData(skin_final, vars.to.regress = "percent.mt")

#PCA
skin_final <- RunPCA(skin_final, npcs = 50, verbose = FALSE)

#Elbow Plot
ElbowPlot(skin_final, ndims=50)

#Choose 35 PCs
```

```{r, message = FALSE}

skin_final <- FindNeighbors(skin_final, reduction = "pca", dims = 1:30)
resolution.range <- seq(from = 0, to = 0.8, by = 0.1)
skin_final <- FindClusters(object = skin_final, resolution = resolution.range)
```


```{r, fig.width=8, fig.height=15}
#Clustree
clustree(skin_final)
```

### Dimplot / Resolutions

```{r, fig.width=12, fig.height=12}
DimPlot(skin_final, reduction = "umap", shuffle = T,
        group.by = c("originalexp_snn_res.0.2", "originalexp_snn_res.0.3", "originalexp_snn_res.0.4", "originalexp_snn_res.0.5"), ncol = 2)

```

### Dimplot / Sample

```{r, fig.width=10, fig.height=6}

DimPlot(skin_final, reduction = "umap", shuffle = T,
        group.by = c("sample"))

#Clearly needs to be re-integrated 
```

```{r, message=F, echo = F, include=F }
skin_final[["originalexp"]] <- split(skin_final[["originalexp"]], f = skin_final$sample)

skin_harmony <- IntegrateLayers(
  object = skin_final, method = HarmonyIntegration, npcs = 25,
  orig.reduction = "pca", new.reduction = "harmony", max.inter.harmony = 10,
  verbose = TRUE
)
```

```{r, message=FALSE, echo = F}

#PCA
skin_final <- RunPCA(skin_final, npcs = 50, verbose = FALSE)

#Elbow Plot
ElbowPlot(skin_final, ndims=50)
skin_harmony <- FindNeighbors(skin_harmony, reduction = "harmony", dims = 1:30)
resolution.range <- seq(from = 0, to = 0.8, by = 0.1)
skin_harmony <- FindClusters(object = skin_harmony, resolution = resolution.range)
clustree(skin_harmony)
```

```{r, message=FALSE}
skin_harmony <- RunUMAP(skin_harmony, reduction = "harmony", dims = 1:30, reduction.name = "umap.harm")
#Rejoin layers
skin_harmony <- JoinLayers(skin_harmony)
```
#### Look at the resolution, UMAP per sample & feature plots

```{r, message=FALSE, fig.height=10, fig.width= 10}
DimPlot(skin_harmony, group.by = c("originalexp_snn_res.0.3", "originalexp_snn_res.0.4", "originalexp_snn_res.0.5", "originalexp_snn_res.0.6"), reduction = "umap.harm")
```

```{r, message=FALSE, fig.height=5, fig.width= 15}
DimPlot(skin_harmony, split.by = "sample", reduction = "umap.harm")
```


```{r, fig.width=10, fig.height=5}

FeaturePlot_scCustom(skin_harmony, reduction = "umap.harm", features = c("percent.mt", "percent.rp", "percent.hb"))

```

### Set Resolution

```{r, fig.width=7, fig.height=7}

DimPlot(skin_harmony, reduction = "umap.harm", label = T)

```

```{r}
#Will set at 0.4 for now and revisit later in case it doesn't make sense
skin_harmony$seurat_clusters <- skin_harmony$originalexp_snn_res.0.4
skin_harmony <- Seurat::SetIdent(skin_harmony, value = skin_harmony$originalexp_snn_res.0.4)
table(skin_harmony$seurat_clusters)
```



## Re-Annotation (https://www.nature.com/articles/s41467-021-24110-y/figures/1)

### Markers
```{r}

Endo <- c("SELE", "PECAM1","VWF")
Fib <- c("COL1A1", "COL1A2", "COL3A1")
SMC <- c("TAGLN", "ACTA2", "TPM2")
KRT <- c("KRT14", "KRT1", "KRT10", "KRT5")
IMM <- c("LYZ", "HLA-DRA")
LYME <- c("COL21", "LYVE1")
NEU <- c("NRXN1")
SGC <- c("SCGB1B2P", "SCGB1D2")
Mela <- c("TYRP1", "PMEL")
Tcell <- c("CCR7", "CD8A", "IL7R", "CD3E")
```

#### Endothelial Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(filtered_seurat_harmony, features = Endo, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = Endo, reduction = "umap.harm")
```


```{r, fig.width=10, fig.height=10}
#VlnPlot_scCustom(skin_harmony, features = Fib, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = Fib, reduction = "umap.harm")
```

```{r, fig.width=10, fig.height=10}
#VlnPlot_scCustom(skin_harmony, features = SMC, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = SMC, reduction = "umap.harm")

```

```{r, fig.width=7, fig.height=7}
#VlnPlot_scCustom(skin_harmony, features = KRT, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = KRT, reduction = "umap.harm")
```
#### Immune Cells 

```{r, fig.width=8, fig.height=5}
#VlnPlot_scCustom(skin_harmony, features = IMM, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = IMM, reduction = "umap.harm")
```
#### Lymphatic Endothelial Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(skin_harmony, features = LYME, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = LYME, reduction = "umap.harm")
```

#### Neuronal Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(skin_harmony, features = NEU, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = NEU, reduction = "umap.harm")
```
#### Sweat Gland Cells 

```{r, fig.width=5, fig.height=5}
#VlnPlot_scCustom(skin_harmony, features = SGC, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = SGC, reduction = "umap.harm")
```

#### Melanocytes

```{r, fig.width=10, fig.height=5}
#VlnPlot_scCustom(skin_harmony, features = Mela, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = Mela, reduction = "umap.harm")
```
#### T Cells 

```{r, fig.width=7, fig.height=7}
#VlnPlot_scCustom(skin_harmony, features = Tcell, pt.size = 0)
FeaturePlot_scCustom(skin_harmony, features = Tcell, reduction = "umap.harm")
```

```{r, fig.width=7, fig.height=7}
#VlnPlot_scCustom(skin_harmony, features = Tcell, pt.size = 0)
cluster4_markers <- FindMarkers(skin_harmony, ident.1 = "14")
head(cluster4_markers)
```



### Annotating the final UMAP 

```{r}

#new.cluster.ids <- c("T-Cells", "Fibroblasts", "Endothelial Cells", "Smooth Muscle Cells", "Immune Cells", "Endothelial Cells",
#                     "Smooth Muscle Cells","Fibroblasts", "Keratinocytes", "Fibroblasts", "T-Cells", "Sweat Gland Cells", "Lymphatic Endothelial #Cells",
#                     "Neuronal Cells","Fibroblasts", "Smooth Muscle Cells", "Melanocytes")

#length(new.cluster.ids)
#length(levels(skin_harmony$seurat_clusters))
```

```{r}
#names(new.cluster.ids) <- levels(skin_harmony$seurat_clusters)
#skin_harmony <- RenameIdents(skin_harmony, new.cluster.ids)

DimPlot(skin_harmony, reduction = "umap.harm", label = TRUE, pt.size = 0.2, repel = T, label.size = 4)


```
```{r}

skin_final <- saveRDS(skin_harmony, 
              here::here(output_dir_data, "skin_final.RDS"))

```
